---
title: "pre_process"
format: html
editor: visual
---

# Load libraries

```{r}
library(Seurat)
library(data.table)
library(ggplot2)
library(dplyr)
library(BiocSingular)
library(scDblFinder)
library(SingleCellExperiment)
library(SingleR)
library(scater)
```

# Load data

```{r}
covid <- readRDS("~/scRNA/data/covid_nbt_main.rds")
```

```{r}
covid <- UpdateSeuratObject(covid)
```

# 1. Explore data and subset

```{r}
covid
```

```{r}
covid[[]]
```

```{r}
unique(covid$severity)
```

```{r}
covid_sub <- subset(covid, subset = severity == "control" | severity == "critical")
covid_sub$severity <- factor(covid_sub$severity)
```

## UMIs per cell vs. Severity

```{r}
covid_sub[[]] |> 
ggplot(aes(color= severity , x= nCount_RNA, fill=severity)) +
geom_density(alpha = 0.2) +
scale_x_log10() +
theme_classic() +
ylab("Cell density") +
geom_vline(xintercept = 500)
```

The UMI counts per cell should generally be above 500

```{r}
covid_sub <- subset(covid_sub, subset = nCount_RNA > 500)
```

# 2. Quality Control

1.  The number of unique genes detected in each cell.

-   low-quality cells or empty droplets will often have very few genes

-   cell doublets or multiplets may exhibit an aberrantly high gene count

2.  The percentage of reads that map to the mitochondrial genome

-   low-quality / dying cells often exhibit extensive mitochondrial contamination

```{r}

covid_sub[["percent.mt"]] <- PercentageFeatureSet(covid_sub, pattern = "^MT-")


covid_sub <- PercentageFeatureSet(covid_sub, "^RP[SL]", col.name = "percent_ribo")
```

```{r}
covid_sub[[]]
```

## Quality metrics

Reset cell annotation

```{r}
Idents(covid_sub) <- "SeuratProject"
```

Visualize quality metrics

```{r}

VlnPlot(covid_sub, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent_ribo"), ncol = 4)
```

### Filter for quality metrics

```{r}
#covid_sub <- subset(covid_sub, 
                    subset = nFeature_RNA > ???? &
                      nFeature_RNA < ???? &
                      percent.mt < ???)
```

## Genes constitute most of the UMIs in a single cell

```{r}
par(mar = c(4, 8, 2, 1))
C <- covid_sub@assays$RNA@counts
C <- Matrix::t(Matrix::t(C)/Matrix::colSums(C)) * 100
most_expressed <- order(apply(C, 1, median), decreasing = T)[20:1]

selected_genes <- t(as.matrix(C[most_expressed, ]))

boxplot(as.matrix(t(C[most_expressed, ])), cex = 0.1, las = 1, xlab = "% total count per cell",
  col = (scales::hue_pal())(20)[20:1], horizontal = TRUE)
```

### Filter for genes

```{r}
#covid_sub <- covid_sub[ ! grepl('^RP[SL]', rownames(pbmc.MTB)), ]
```

# 3. Normalization

```{r}
covid_sub <- NormalizeData(covid_sub, normalization.method = "LogNormalize", scale.factor = 10000)
```

# 4. Identify Highly Variable Features

```{r}
covid_sub <- FindVariableFeatures(covid_sub, selection.method = "vst", nfeatures = 2000)
```

```{r}

top10 <- head(VariableFeatures(covid_sub), 10)

plot1 <- VariableFeaturePlot(covid_sub)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

# 5. Scale data

```{r}
all.genes <- rownames(covid_sub)
covid_sub <- ScaleData(covid_sub, features = all.genes)
```

# 6. Visualization

```{r}
covid_sub <- RunPCA(covid_sub, features = VariableFeatures(object = covid_sub))
```

```{r}
ElbowPlot(covid_sub) 
```

```{r}
covid_sub <- FindNeighbors(covid_sub, dims = 1:15)
covid_sub<- FindClusters(covid_sub, resolution = c(0.3, 0.5, 0.7))
```

```{r}
covid_sub <- RunUMAP(covid_sub, dims = 1:15)
p1<- DimPlot(covid_sub, reduction = "umap")

DimPlot(covid_sub, group.by = "RNA_snn_res.0.3", label = TRUE)
```

```{r}
Idents(covid_sub) <- "RNA_snn_res.0.5"
```

# 7. Doublets

```{r}

sce <- as.SingleCellExperiment(covid_sub)

top.var <- VariableFeatures(covid_sub)
```

```{r}
dbl.dens <- computeDoubletDensity(sce, subset.row=top.var, 
                                  d=ncol(reducedDim(sce)))
summary(dbl.dens)
```

```{r}
sce$DoubletScore <- dbl.dens
plotUMAP(sce, colour_by="DoubletScore")
```

```{r}
dbl.calls <- doubletThresholding(data.frame(score=dbl.dens),
                                 method="griffiths", returnType="call")
summary(dbl.calls)
```

# SAVE OBJECT

```{r}
saveRDS(covid_sub, "covid_sub")
```
